Scaling through abstractions – high-performance vectorial wave simulations for seismic inversion with Devito

### 2020-06-02 response.md

Dear Editor,

First of all, we would like to thank the reviewers for their work and constructive feedback. Below, we address all of their questions and observations.

## Review of pap429s2 by Reviewer 1

> If you run your model with a variable number of cores, how is the weak and strong scaling? Can you comment on that? Further, you mention that you validate performance against a wave propagator - how about validation of simulation quality? How is that done?

At the time of the experimentation, we ran into two main issues:
1. Limited resources on Azure, including the access to an Infiniband-enabled system.
2. Most of the MPI-level optimizations that are today in Devito still needed tuning and consolidation.

We subsequently managed to work around point one above. Initial scaling experiments have been published in [..., page ..., figure ...], although this work still doesn't exploit the latest advancements in Devito v4.2 (e.g., computation/communication overlap). Our manuscript centers on the development of HPC production-grade wave propagators from a high-level language, and while performance is one of the cornerstones of our work, showing off Devito at extreme scale was not an objective. We have preliminary evidence that the latest Devito v4.2 has remarkable parallel efficiency, but we felt that publishing these results today would be premature -- we at least need testing and reproducibility on several architectures and applications.

About validation of simulation quality -- we did not include it in this paper as we think it is outside of the scope of the presented results, but the simulation quality is verified in two ways. As described in [@devito-api], the numerical accuracy is compared to the analytical solution for the acoustic wave equation as well as the convergence rate of the finite-difference discretization. The verification of TTI is not as simple as we do not have an analytical solution; we will be using the method of manufactured solution (MMS), but this is still work in progress. However, all of the key parts, such as finite-difference accuracy and comparison with isotropic results for zero-ed anisotropic parameters are implemented and part of our continuous integration system.

## Review of pap429s2 by Reviewer 2

@@

* However, it is not very suitable for SC

As partly explained in our reply to R1, we conceived this work as an application-oriented paper based on high-level abstractions and compilation technology. We're in the middle of an epochal change for computational science, with legacy codes being superseded, gradually, by DSLs; our work is just yet another small step in this direction. However, unlike many other works centered on DSLs, ours builds on -- and contributes to -- what is nowadays an established framework for seismic imaging. We believe that this should be of strong interest for the SC community, especially since -- we believe -- the methodology and philosophy underpinning Devito is reusable in computational science.

@@

*  Parallelization is pretty straightforward, uniform decomposition of relatively straightforward operation.

We remark that our work does _not_ describe some ad-hoc MPI-based domain decomposition; what Devito can do is _generation_ of MPI parallel code from a mathematical language for solving partial differential equations. This requires considerable software engineering (especially for an open source project such as Devito that was born in academia): we're talking about thousands of lines of carefully engineered code, plus tests and examples, to enable users to just type `mpirun -n X python ...` and get their code to run on a distributed-memory system with virtually no changes to their code. Secondly, as we explained in the paper [section ...], it's not just "classic domain decomposition": Devito orchestrates sparse functions, implements distributed numpy arrays, and performs several compiler-level optimazions for effective halo exchange. Based on our experience, this is a fairly sophisticated system, quite far from being "straightforward".

@@

* The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization

High arithmetic intensity does not imply "non challenging optimization". In fact, it's the exact opposite, and there are several examples in computational science that we could use as evidence, even outside of seismic. Take high-order finite (spectral) elements for instance -- you may get an extremely good arithmetic intensity, but without tensor-product basis functions and suitable optimizations such as sum-factorization you'll likely end up with a notable, yet totally misleasing, GFlops/s performance, as a huge fraction of flops will stem from redundant computation (e.g., common sub-expressions, lack of code hoisting, distributivity). The same story happens with our TTI. As we explain in the paper [section ...], without capturing the cross-iteration common sub-expressions (due to high order derivatives), we would have obtained remarkable GFlops/s performance, but terrible GPoints/s. And capturing such redundancies is indeed quite challenging. In fact, this is one of the reasons several companies are today so much interested in Devito: machines (compilers) can be better than humans at detecting redundancies, and definitely way faster than them.

### 2020-06-02 response.md

## Review of pap429s2 by Reviewer 2

> However, it is not very suitable for SC

As partly explained in our reply to R1, we conceived this work as an application-oriented paper based on high-level abstractions and compilation technology. We're in the middle of an epochal change for computational science, with legacy codes being superseded, gradually, by DSLs; our work is just yet another small step in this direction. However, unlike many other works centered on DSLs, ours builds on -- and contributes to -- what is nowadays an established framework for seismic imaging. We believe that this should be of strong interest for the SC community, especially since -- we believe -- the methodology and philosophy underpinning Devito is reusable in computational science.

> Parallelization is pretty straightforward, uniform decomposition of relatively straightforward operation.

@@

We remark that our work does _not_ describe some ad-hoc MPI-based domain decomposition; what Devito can do is _generation_ of MPI parallel code from a mathematical language for solving partial differential equations. This requires considerable software engineering (especially for an open source project such as Devito that was born in academia): we're talking about thousands of lines of carefully engineered code, plus tests and examples, to enable users to just type `mpirun -n X python ...` and get their code to run on a distributed-memory system with virtually no changes to their code. Secondly, as we explained in the paper [section ...], it's not just "classic domain decomposition": Devito orchestrates sparse functions, implements distributed numpy arrays, and performs several compiler-level optimazions for effective halo exchange. Based on our experience, this is a fairly sophisticated system, quite far from being "straightforward".

> The computational kernels have really high arithmetic intensity that is great for the domain, but not as challenging for optimization

@@

High arithmetic intensity does not imply "non challenging optimization". In fact, it's the exact opposite, and there are several examples in computational science that we could use as evidence, even outside of seismic. Take high-order finite (spectral) elements for instance -- you may get an extremely good arithmetic intensity, but without tensor-product basis functions and suitable optimizations such as sum-factorization you'll likely end up with a notable, yet totally misleasing, GFlops/s performance, as a huge fraction of flops will stem from redundant computation (e.g., common sub-expressions, lack of code hoisting, distributivity). The same story happens with our TTI. As we explain in the paper [section ...], without capturing the cross-iteration common sub-expressions (due to high order derivatives), we would have obtained remarkable GFlops/s performance, but terrible GPoints/s. And capturing such redundancies is indeed quite challenging. In fact, this is one of the reasons several companies are today so much interested in Devito: machines (compilers) can be better than humans at detecting redundancies, and definitely way faster than them.

> The parallelism applied so far is straight up MPI which is a bit behind times these days.

### 2020-06-02 response.md

### Reply

> One question that arises, in the performance is how does Devito account for FMA (Fused Multiple Adds)?

@@

The operations determining the performance of a code are the source level ones, while FMA is a lower level concept (ie, a scalar FMA would account for two operations). This is not Devito, but rather then model adopted everywhere, otherwise two performance values from two different architectures, one with FMAs the other without, would not even be comparable.

> Scalability performance would also be interesting.

See reply to R1 -- basically, (i) lack of resources and (ii) Devito not ready yet for extreme scale at the time we ran the experiments.

> It appears much of the performance result are provided in the references so it is unclear as to what the purpose of this paper is for an SC audience.

We are showing performance results of three different codes -- TTI, elastic, and isotropic acoustic -- thus exploring different physics and discretizations. These results are, respectively, in sections [...]. As we explain in our reply to R2, ours is an application-oriented paper built off high-level abstractions running on a cloud-based system. To achieve this, we're using several bleeding edge technologies -- an established DSL-based system such as Devito (used in companies and academia), the Azure cloud, Docker (for ease of installation, use and reproducibility), the rapidly-spreading GitHub Actions for continuos integration, and much more. The use and orchestration of these technologies is thouroughly described in the paper, and this should be, in our opinion, of strong interest for the SC audience.

> For actual seismic surveys, the imaging (interpretation) part is only 10% of the survey. Of course, one does us the RTM in developing the velocity model which adds to complexity in a way. Ease of use coupled with excellent scalable performance while important doesn't address the real "industrial" problems, it helps a little.

The industrial problem does indeed involve muliple parts, such as data acquisition, processing, and, most importantly, velocity model building. We do not claim here to solve all these problems; however, we demonstrate that the combination of advanced technolgies such as high-level DSLs and compilers is key to highly efficient sesimic modeling and imaging at industrial scale. The complexity of model building requires advanced optimization methods. The straightforward access to computationally and easy-to-develop wave-equation solvers enable research and development for seismic inversion. Devito is therfore used for the development of inital background model building [@mojica2019tab] or cycle skipping mitigating methods [@rizzuti2020EAGEtwri; @louboutin2020SEGtwri] and compressive least square migration [@witte2018cls] based on a linear algebra framework for seismic inversion built on top of devito [@witte2018alf]. We have added a clarification in the introduction.

### 2020-06-02 paper.tex
Typos

We here provide a succinct description of distributed-memory parallelism
in Devito; the interested reader should refer to the MPI tutorial at
\href{https://github.com/devitocodes/devito/blob/v4.2/examples/mpi/overview.ipynb}{mpi-notebook}
for thorough explanations and practical examples.

@@

Earlier performance benchmarks mainly focused on roofline model analysis
\textcolor{red}{\cite{sevendwarfs, patterson, williams2009roofline, louboutin2016ppf}}.
In this study, for completeness, the runtime of \devito is therefore
compared to that of the open source hand-coded propagator
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc}. This propagator,
described in \cite{thorbecke}, is a state of the art elastic kernel
(Equation~\ref{elas1}) and the comparisons presented here were carried out
in collaboration with its author.
\textcolor{red}{It should be emphasized that direct model comparison in
the area of exploration geophysics is challenging. The majority of production
codes are either proprioty or are simply not publicly available. Further,
\devito should not be compared against pure stencil frameworks,
such as YASK \cite{yask}, since they are unable to compute the sparse operations
required for source injection and receiver interpolation (and are therefore not
suitable for use in measurement based inversion problems). The comparison carried out
here is conducted in two dimensions. Whilst large scale 3D scaling comparisons are theoretically
possible, compute time on the Cloud is expensive. Such a comparison is not feasible at this
time and will therefore be left for future work.}

>>>>

We here provide a succinct description of distributed-memory parallelism
in Devito; the interested reader should refer to the MPI tutorial at
\href{https://github.com/devitocodes/devito/blob/v4.2/examples/mpi/overview.ipynb}{mpi-notebook}
for thorough explanations and practical examples.
\textcolor{red}{We highlight the main features of the distributed framework tha enable the industry scale
applications we describe in this paper. A more thorough study of its implementation, including `mpi4py' and
a complete scaling analysis will be presented in a later paper. Initial results associated with early version of \devito's
distributed computing capabilities can be found in \cite{luporini2019adp} on a conventional cluster and 
in \cite{witte2019ecl} that contains results on AWS.}

@@

Earlier performance benchmarks mainly focused on roofline model analysis
\textcolor{red}{\cite{sevendwarfs, patterson, williams2009roofline, louboutin2016ppf}}.
In this study, for completeness, the runtime of \devito is therefore
compared to that of the open source hand-coded propagator
\href{https://github.com/JanThorbecke/OpenSource.git}{fdelmodc}. This propagator,
described in \cite{thorbecke}, is a state of the art elastic kernel
(Equation~\ref{elas1}) and the comparisons presented here were carried out
in collaboration with its author.
\textcolor{red}{It should be emphasized that direct model comparison in
the area of exploration geophysics is challenging. The majority of production
codes are either proprietary or are simply not publicly available. Further,
\devito should not be compared against pure stencil frameworks,
such as YASK \cite{yask}, since they are unable to compute the sparse operations
required for source injection and receiver interpolation (and are therefore not
suitable for use in measurement based inversion problems). The comparison carried out
here is conducted in two dimensions. Whilst large scale 3D scaling comparisons are theoretically
possible, compute time on the Cloud is expensive. Such a comparison is not feasible at this
time and will therefore be left for future work.}
